\TheTitle

### 2019-07-14 Paper1/introduction.tex
started correcting Martins comments
smaller stuff fixed

Numerical methods for approximating the moment $\bm{\hat u}_i$ can be divided into intrusive and non-intrusive techniques. A popular non-intrusive method is the stochastic-Collocation (SC) method, see e.g. \cite{xiu2005high,babuvska2007stochastic,loeven2008probabilistic}, which computes the moments with the help of a numerical quadrature rule: For a given set of $Q$ quadrature weights $w_k$ and quadrature points $\bm{\xi}_k$, the moments are approximated by
\begin{align*}
\bm{\hat u}_i = \langle \bm{u}\varphi_i \rangle \approx \sum_{k = 1}^{Q}w_k \bm{u}({t,\bm{x},\bm{\xi}_k})\varphi_i(\bm{\xi}_k)f_{\Xi}(\bm{\xi}_k).
\end{align*} 
Since the solution at a fixed quadrature point can be computed by a standard deterministic solver, the SC method does not require a significant implementation effort. Furthermore, SC is embarrassingly parallel, since the required computations can be carried out in parallel on different cores. A downside of collocation methods are aliasing effects, which stem from the inexact approximation of integrals. Furthermore, collocation methods require more runs of the deterministic solver than intrusive methods \cite{xiu2009fast,alekseev2011estimation}. Despite their easy implementation, collocation methods face challenges when examining unsteady problems, since in this case, the solution needs to be written out, i.e. stored in an external file, at all quadrature points in every time step to compute the time evolution of the moments. This contradicts modern HPC paradigms, which aim at reducing the amount of data produced by numerical methods.

@@

Corresponding to the definition of $\bm{\hat u}$, we collect the $N$ basis function for which $\vert i \vert \leq M$ holds in $\bm{\varphi}:=(\varphi_i)_{|i|\leq M}\in\mathbb{R}^{N+1}$. When using the stochastic-Galerkin method to close \eqref{eq:nonClosedMomentSystem}, the resulting moment system is not necessarily hyperbolic \cite{poette2009uncertainty} and the solution needs to be manipulated \cite{schlachter2018hyperbolicity} in order to prevent a failure of the method. A generalization of stochastic-Galerkin, which ensures hyperbolicity is the Intrusive Polynomial Moment (IPM) closure \cite{poette2009uncertainty}: The closure is given by a constraint optimization problem. For a given convex entropy $s:\mathbb{R}^p\to\mathbb{R}$ for the original problem \eqref{eq:hyperbolicProblem}, this optimization problem is given by

@@

Rewritten in its dual form, \eqref{eq:primalProblem} is transformed into an unconstraint optimization problem. Defining the variables $\bm{\lambda}_i\in\mathbb{R}^p$ where $i$ is again a multi index, gives the unconstrained dual problem

@@

i.e. the IPM method yields a physically correct entropy solution. This again underlines a weakness of stochastic-Galerkin: If $s(\bm{u}) = \frac{1}{2}\bm{u}^T\bm{u}$ is not a correct entropy of the original problem, the SG method can lead to non-physical solution values, which can then cause a failure of the method. The main weakness of the IPM method is its run time, since it requires the repeated evaluation of \eqref{eq:ansatz}, which involves solving the optimization problem \eqref{eq:dualProblem}. Hence, the desirable properties of IPM come along with significantly increased numerical costs. However, IPM and minimal entropy methods in general are well suited for modern HPC architecture, which can be used to reduce the run time \cite{garrett2015optimization}.

>>>>

Numerical methods for approximating the moment $\bm{\hat u}_i$ can be divided into intrusive and non-intrusive techniques. A popular non-intrusive method is the stochastic-Collocation (SC) method, see e.g. \cite{xiu2005high,babuvska2007stochastic,loeven2008probabilistic}, which computes the moments with the help of a numerical quadrature rule: For a given set of $Q$ quadrature weights $w_k$ and quadrature points $\bm{\xi}_k$, the moments are approximated by
\begin{align*}
\bm{\hat u}_i = \langle \bm{u}\varphi_i \rangle \approx \sum_{k = 1}^{Q}w_k \bm{u}({t,\bm{x},\bm{\xi}_k})\varphi_i(\bm{\xi}_k)f_{\Xi}(\bm{\xi}_k).
\end{align*} 
Commonly, SC uses sparse grids as quadrature rule, since they posses a reduced number of collocation points for multi-dimensional problems. Compared to tensorized quadrature sets, which require $O(M^s)$ quadrature points, sparse grids use $Q = O(M(\log_2(M)^{s-1}))$ quadrature points to integrate polynomials of total degree $M$ exactly. 
Since the solution at a fixed quadrature point can be computed by a standard deterministic solver, the SC method does not require a significant implementation effort. Furthermore, SC is embarrassingly parallel, since the required computations can be carried out in parallel on different cores. A downside of collocation methods are aliasing effects, which stem from the inexact approximation of integrals. Therefore, collocation methods typically require more runs of the deterministic solver than intrusive methods \cite{xiu2009fast,alekseev2011estimation} to reach a given accuracy. Despite their easy implementation, collocation methods face challenges when examining unsteady problems, since in this case, the solution needs to be written out, i.e. stored in an external file, at all quadrature points in every time step to compute the time evolution of the moments. This contradicts modern HPC paradigms, which aim at reducing the amount of data produced by numerical methods.

@@

Corresponding to the definition of $\bm{\hat u}$, we collect the $N$ basis function for which $\vert i \vert \leq M$ holds in $\bm{\varphi}:=(\varphi_i)_{|i|\leq M}\in\mathbb{R}^{N+1}$. When using the stochastic-Galerkin method to close \eqref{eq:nonClosedMomentSystem}, the resulting moment system is not necessarily hyperbolic \cite{poette2009uncertainty} and the solution needs to be manipulated \cite{schlachter2018hyperbolicity} in order to prevent a failure of the method. A generalization of stochastic-Galerkin, which ensures hyperbolicity is the Intrusive Polynomial Moment (IPM) closure \cite{poette2009uncertainty}: The closure is given by a constrained optimization problem. For a given convex entropy $s:\mathbb{R}^p\to\mathbb{R}$ for the original problem \eqref{eq:hyperbolicProblem}, this optimization problem is given by

@@

Rewritten in its dual form, \eqref{eq:primalProblem} is transformed into an unconstrained optimization problem. Defining the variables $\bm{\lambda}_i\in\mathbb{R}^p$ where $i$ is again a multi index, gives the unconstrained dual problem

@@

i.e. the IPM method yields a physically correct entropy solution. This again underlines a weakness of stochastic-Galerkin: If $s(\bm{u}) = \frac{1}{2}\bm{u}^T\bm{u}$ is not a correct entropy of the original problem, the SG method can lead to non-physical solution values, which can then cause a failure of the method. The main weakness of the IPM method is its computational effort, since it requires the repeated evaluation of \eqref{eq:ansatz}, which involves solving the optimization problem \eqref{eq:dualProblem}. Hence, the desirable properties of IPM come along with significantly increased numerical costs. However, IPM and minimal entropy methods in general are well suited for modern HPC architecture, which can be used to reduce the run time \cite{garrett2015optimization}.

### 2019-07-14 Paper1/main.tex
paper continued
results started + details IPM added
started correcting Martins comments
smaller stuff fixed

% Declare title and authors, without \thanks
\newcommand{\TheTitle}{A semi-intrusive Code Framework for Uncertainty Quantification}

@@

\begin{abstract}
Methods for quantifying the effects of uncertainties in hyperbolic problems can be divided into intrusive and non-intrusive techniques. A main drawback of intrusive methods is the need to implement new code, whereas collocation methods recycle a given deterministic solver. Furthermore, intrusive methods come with increased computational costs, making it hard to compete with non-intrusive methods.
In this work, we present a code framework, which facilitates the implementation of intrusive methods by recycling parts of deterministic codes. Additionally, we introduce methods to decrease numerical costs of intrusive methods for steady problems. We demonstrate the effectiveness of the proposed strategies by comparing results of the uncertain NACA0012 testcase as well as a bent shock tube experiment.
\end{abstract}

@@

\input{introduction}
\input{framework}
\section{Strategies for steady problems}
\input{colloc_ipm}
\input{one_shot_ipm}
\input{adaptivity}
\input{results}
\input{summary_outlook}

>>>>

% Declare title and authors, without \thanks
\newcommand{\TheTitle}{Acceleration strategies for intrusive methods in Uncertainty Quantification}

@@

\begin{abstract}
Methods for quantifying the effects of uncertainties in hyperbolic problems can be divided into intrusive and non-intrusive techniques. A main drawback of intrusive methods is the need to implement new code, whereas collocation methods recycle a given deterministic solver. Moreover, non-intrusive methods are embarrassingly parallel and therefore easy to parallelize.
In this work, we present a code framework, which facilitates the implementation of intrusive methods by recycling parts of deterministic codes. Additionally, we introduce methods to decrease numerical costs of intrusive methods for steady problems. We demonstrate the effectiveness of the proposed strategies by comparing results of the uncertain NACA0012 testcase as well as a bent shock tube experiment.
\end{abstract}

@@

\input{introduction}
\input{framework}
\section{Strategies for steady problems}
\input{colloc_ipm}
\input{one_shot_ipm}
\input{adaptivity}
\input{results}
\input{summary_outlook}
\input{appendix}

### 2019-07-14 Paper1/one_shot_ipm.tex
started correcting Martins comments
smaller stuff fixed

In the following, we will show that this iteration converges, if the chosen initial condition is sufficiently close to the steady state solution. For this we take an approach commonly chosen to prove local convergence properties of Newton's method: In Theorem \ref{th:Contractive}, we show that the iteration function is contractive at its fix point and conclude in Theorem \ref{th:localConvergence} that this yields local convergence:

@@

Assume that the classical IPM iteration is contractive at its fix point $\bm{\hat u}^*$. Then the Jacobi matrix $\bm{J}$ of the One-Shot IPM iteration \eqref{eq:oneshotIPM} has a spectral radius $\rho(\bm{J})<1$ at the fix point $(\bm{\lambda}^*,\bm{\hat u}^*)$ if the preconditioner $\bm{B} = \langle \nabla \bm{u}_s(\bm{\varphi}^T\bm{\lambda}_j^{n})\bm{\varphi}\bm{\varphi}^T\rangle^{-1}$, i.e. the Hessian of the dual problem is used.

@@

Since we assume that the classical IPM scheme is contractive at its fix point, we have $\rho (\nabla_{\bm{\hat u}}\bm{\tilde c}(\bm{\hat u}^*))<1$ with $\nabla_{\bm{\hat u}}\bm{\tilde c}\in\mathbb{R}^{N\cdot N_x\times N\cdot N_x}$ defined by

@@

Now we begin to derive the spectrum of the \textit{One-Shot IPM} iteration \eqref{eq:oneshotIPM}. Note that in its current form this iteration is not really a fix point iteration, since it uses the time updated dual variables in \eqref{eq:oneshotIPMmoment}. To obtain a fix point iteration, we plug the dual iteration step \eqref{eq:oneshotIPMdual} into the moment iteration \eqref{eq:oneshotIPMmoment} to obtain

@@

where we used $\bm{B} = \langle \nabla \bm{u}_s(\bm{\varphi}^T\bm{\lambda}_j^n)\bm{\varphi}\bm{\varphi}^T \rangle^{-T}$. Recall that at the fix point $(\bm{\lambda}^*,\bm{\hat u}^*)$, we have $\langle \bm{u}_s(\bm{\varphi}^T\bm{\lambda}_j^n)\bm{\varphi}^T \rangle^T = \bm{\hat u}$, hence one obtains $\partial_{\bm{\lambda}} \bm{d}=\bm{0}$. For the block $\partial_{\bm{\hat u}} \bm{d}$, we get 

@@

by the choice of $\bm{B}$ as well as \eqref{eq:cTildeDer} and \eqref{eq:dudlambdaex}. We obtain an analogous result for the second and third input. Hence, we have that $\partial_{\bm{\hat u}} \bm{c} = \nabla_{\bm{\hat u}}\bm{\tilde c}$, which only has eigenvalues between $-1$ and $1$ by the assumption that the classical IPM iteration is contractive. Since $\bm{J}$ is an upper triangluar block matrix, the eigenvalues are given by $\lambda\left(\partial_{\bm{\lambda}} \bm{d}\right) = 0$ and $\lambda\left(\partial_{\bm{\hat u}} \bm{c}\right)\in(-1,1)$, hence the One-Shot IPM is contractive around its fix point.

@@

By Theorem \ref{th:Contractive}, the One-Shot scheme is contractive at its fix point. Since we assumed convergence of the classical IPM scheme, we can conclude that all entries in the Jacobian $\bm{J}$ are continuous functions. Furthermore, the determinant of $\bm{\tilde{J}}:=\bm{J}-\lambda \bm{I}_d$ is a polynomial of continuous functions, since

@@

Since the roots of a polynomial vary continuously with its coefficients, the eigenvalues of $\bm{J}$ are continuous w.r.t $(\bm{\lambda},\bm{\hat u})$. Hence there exists an open ball with radius $\delta$ around the fix point in which the eigenvalues remain in the interval $(-1,1)$.

>>>>

In the following, we will show that this iteration converges, if the chosen initial condition is sufficiently close to the steady state solution. For this we take an approach commonly chosen to prove local convergence properties of Newton's method: In Theorem \ref{th:Contractive}, we show that the iteration function is contractive at its fixed point and conclude in Theorem \ref{th:localConvergence} that this yields local convergence:

@@

Assume that the classical IPM iteration is contractive at its fixed point $\bm{\hat u}^*$. Then the Jacobi matrix $\bm{J}$ of the One-Shot IPM iteration \eqref{eq:oneshotIPM} has a spectral radius $\rho(\bm{J})<1$ at the fixed point $(\bm{\lambda}^*,\bm{\hat u}^*)$ if the preconditioner $\bm{B} = \langle \nabla \bm{u}_s(\bm{\varphi}^T\bm{\lambda}_j^{n})\bm{\varphi}\bm{\varphi}^T\rangle^{-1}$, i.e. the Hessian of the dual problem is used.

@@

Since we assume that the classical IPM scheme is contractive at its fixed point, we have $\rho (\nabla_{\bm{\hat u}}\bm{\tilde c}(\bm{\hat u}^*))<1$ with $\nabla_{\bm{\hat u}}\bm{\tilde c}\in\mathbb{R}^{N\cdot N_x\times N\cdot N_x}$ defined by

@@

Now we begin to derive the spectrum of the \textit{One-Shot IPM} iteration \eqref{eq:oneshotIPM}. Note that in its current form this iteration is not really a fixed point iteration, since it uses the time updated dual variables in \eqref{eq:oneshotIPMmoment}. To obtain a fixed point iteration, we plug the dual iteration step \eqref{eq:oneshotIPMdual} into the moment iteration \eqref{eq:oneshotIPMmoment} to obtain

@@

where we used $\bm{B} = \langle \nabla \bm{u}_s(\bm{\varphi}^T\bm{\lambda}_j^n)\bm{\varphi}\bm{\varphi}^T \rangle^{-T}$. Recall that at the fixed point $(\bm{\lambda}^*,\bm{\hat u}^*)$, we have $\langle \bm{u}_s(\bm{\varphi}^T\bm{\lambda}_j^n)\bm{\varphi}^T \rangle^T = \bm{\hat u}$, hence one obtains $\partial_{\bm{\lambda}} \bm{d}=\bm{0}$. For the block $\partial_{\bm{\hat u}} \bm{d}$, we get 

@@

by the choice of $\bm{B}$ as well as \eqref{eq:cTildeDer} and \eqref{eq:dudlambdaex}. We obtain an analogous result for the second and third input. Hence, we have that $\partial_{\bm{\hat u}} \bm{c} = \nabla_{\bm{\hat u}}\bm{\tilde c}$, which only has eigenvalues between $-1$ and $1$ by the assumption that the classical IPM iteration is contractive. Since $\bm{J}$ is an upper triangluar block matrix, the eigenvalues are given by $\lambda\left(\partial_{\bm{\lambda}} \bm{d}\right) = 0$ and $\lambda\left(\partial_{\bm{\hat u}} \bm{c}\right)\in(-1,1)$, hence the One-Shot IPM is contractive around its fixed point.

@@

By Theorem \ref{th:Contractive}, the One-Shot scheme is contractive at its fixed point. Since we assumed convergence of the classical IPM scheme, we can conclude that all entries in the Jacobian $\bm{J}$ are continuous functions. Furthermore, the determinant of $\bm{\tilde{J}}:=\bm{J}-\lambda \bm{I}_d$ is a polynomial of continuous functions, since

@@

Since the roots of a polynomial vary continuously with its coefficients, the eigenvalues of $\bm{J}$ are continuous w.r.t $(\bm{\lambda},\bm{\hat u})$. Hence there exists an open ball with radius $\delta$ around the fixed point in which the eigenvalues remain in the interval $(-1,1)$.
